Robust Quantum Optimal Control

### 2020-09-07 paper/paper.tex
updated figures and abstract

%% ABSTRACT
\begin{abstract}
  %% - 1 + 2 topic importance
  %% - 1 what I have done
  %% - few sentences about primary results
  The ability to engineer high fidelity operations on quantum processors in the presence of
  systematic errors and decoherence remains the primary challenge requisite to achieving quantum advantage.
  Quantum optimal control (QOC) techniques have proven effective in realizing high fidelity operations,
  but they require exquisite calibration to be performant. In this work we employ robust control techniques
  to achieve high fidelity gates in the presence of parameter spreads up to 5$\%$.
  Additionally, current QOC techniques mitigate decoherence effects due to longitudinal relaxation by integrating
  Lindblad master equation dynamics, which drastically increase the computational complexity
  of the optimization algorithm.
  In this work we achieve a factor of $4$ increase in longitudinal relaxation times
  for the gate set of a fluxonium qubit
  using a computationally efficient metric in optimizaiton.

  %% The ability to engineer high fidelity operations on quantum processors in the presence of
  %% systematic errors and decoherence remains the primary challenge requisite to achieving quantum advantage.
  %% Quantum optimal control (QOC) techniques have been developed to mitigate deleterious effects
  %% by manipulating the time-dependent parameters controlling the system. These techniques often
  %% target a single error channel and may be incompatible with experimental constraints. 
  %% In this work, we develop techniques to simultaneously mitigate
  %% longitudinal relaxation and pure dephasing in a framework that can handle arbitrary
  %% experimental constraints.
  %% Optimizing a gate set for a fluxonium qubit as an example, we achieve a factor of 4 increase in
  %% longitudinal relaxation times and outperform
  %% existing dynamic decoupling methods in mitigating pure dephasing due to flux noise, achieving
  %% high fidelity gates in the presence of parameter spreads up to $5\%$.
\end{abstract}

@@

%% F1
\begin{figure}[ht]
  \begin{subfigure}{.6\linewidth}
    \includegraphics[width=\linewidth]{assets/f1a.png}
  \end{subfigure}%
    \begin{subfigure}{.4\linewidth}
    \includegraphics[width=\linewidth]{assets/f1b.png}
  \end{subfigure}

  \begin{subfigure}{\linewidth}
    \includegraphics[width=\linewidth]{assets/f1c.png}
  \end{subfigure}
  \caption{
    (a) $T_{1}$ optimized gates (red) and analytic gates (blue).
    (b) $T_{1}$ interpolation function used in optimization
    and the corresponding $T_{1}$ times for the average absolute amplitude of each gate.
    (c) Master equation simulation with $T_{1}$ dissipation
    for all gates, over the course of concatenated gate applications.
  }
\end{figure}

%% S3
\section{QOC on the Fluxonium}
(Fluxonium Device) In the following we study
the quantum optimal control problem on the fluxonium qubit.
In the two-level
approximation the system Hamiltonian takes the form
\label{eq:hamiltonian}
\begin{align}
  H/h &= \omega_{q} \frac{\sigma_{z}}{2} + a(t) \frac{\sigma_{x}}{2}
\end{align}
where $\omega_{q}$ is the qubit frequency at the flux frustration point,
$a$ is the flux drive amplitude, $h$ is Planck's constant, and $\sigma_{x}, \sigma_{y}$
are Pauli matrices. The flux amplitude $a$ is experimentally
realized by modulating the flux 
threading the device ($\Phi_{\textrm{ext}}$).
The flux amplitude is obtained from the external flux by
$a = 4 \pi \braket{g \lvert \hat{\phi} \rvert e} \rvert_{0.5 \Phi_{0}} E_{L}
\delta \Phi_{\textrm{ext}} / (h \Phi_{0})$
where $\hat{\phi}$ is the phase operator, $E_{L}$ is the characteristic inductance energy, $\Phi_{0}$
is a flux quantum, and
$\delta \Phi_{\textrm{ext}} = \Phi_{\textrm{ext}} - 0.5 \Phi_{0}$ is the flux
offset from the flux frustration point.

@@

%% T1
%% TODO: maybe for appendix.
\begin{table}[ht]
  \begin{tabular}{c | c | c | c}
    & Analytic  & QOC &\\
    Gate & $P_{1}\ (10^{-5})$ & $P_{1}\ (10^{-5})$ & Ratio\\
    \hline
    Z/2 & 5.745 & 1.149 & 5.000\\
    Y/2 & 5.253 & 1.157 & 4.540\\
    X/2 & 16.251 & 6.946 & 2.340\\
  \end{tabular}
  \caption{Probability of longitudinal relaxation for each gate
    evaluated at the gate's duration.}
\end{table}

%% F2
\begin{figure}[ht]
  \begin{subfigure}{.6\linewidth}
    \includegraphics[width=\linewidth]{assets/f2a.png}
  \end{subfigure}%
  \begin{subfigure}{.4\linewidth}
    \includegraphics[width=\linewidth]{assets/f2b.png}
  \end{subfigure}

@@

  \caption{\centering (a) $X/2$ gates robust to qubit frequency offsets
    constructed using the analytic, (2, 4)-point
    sampling, and (2, 3)-derivative methods at a gate duration of $56.8 (ns)$.
    (b) Gate error as a function of qubit frequency detuning from the nominal frequency.
    (c) Average gate error at one standard deviation from the nominal qubit frequency as
    a function fo the gate duration.
  }
\end{figure}

@@

  \caption{\centering
    (a) $X/2$ gates robust to flux offsets constructed using the analytic,
    (2, 4)-point sampling, and (2, 3)-derivative methods at a gate duration
    of $56.8 (ns)$. (b) Mean gate error for 10 samples of 1/$f$ flux noise over 5000
    gate applications.
    %% (c) Stochastic 1/$f$ flux noise.
  }
\end{figure}

%% S5
\section{Robustness to Parameter Deviations}
(Strategy) For the fluxonium, pure dephasing acts
along the flux axis. Low frequency 1/$f$ noise,
Johnson-Nyquist current noise present in the flux bias
lines, and temperature dependent gain fluctuations
may cause the flux amplitude to deviate from its
desired value by an amount $\delta a$. Additionally,
residual calibration errors and fluctuations may cause
the qubit frequency to deviate from its measured value
by an amount $\delta \omega_{q}$.

@@

The second scheme we analyze is the sampling method,
which is well studied in the robust control literature
\cite{manchester2016derivative} \cite{tronarp2016sigma}.
We add additional
states that follow deviant dynamics to the augmented
state vector. The parameters of interest
are altered by a small deviation, typically chosen
to represent the distribution of the parameter.
For example, in the 2-point sampling method we
propagate $\ket{\psi^{\pm, I}(t)}$ where
$\omega_{q}^{\pm} = \omega_{q} \pm \sigma_{\omega_{q}}$,
$\omega_{q}^{I} = \omega_{q}$.
The difference between the final states and the target
state is penalized, resulting in quadratic costs at
each knot point
${\lvert \ket{\psi^{\pm, I}_{k}} - \ket{\psi_{f}} \rvert}^{2}$

(Results) We compare the robust control
schemes to a dynamic decoupling pulse
for a single $R_{x}(\pi/2)$ gate
subject to qubit frequency detuning
($\omega_{q} \gets \omega_{q} + \delta \omega_{q}$)
to demonstrate the applicability of our techniques to mitigate
system parameter deviations.
We use a compensating for off-resonance error
(CORPSE) dynamic decoupling pulse that is optimized
to mitigate first-order error arising due to the detuning.
The pulse is composed of
three rotations about the $\hat{x}$, $-\hat{x}$, and $\hat{x}$ axes
of the Bloch sphere
designed to compensate for erroneous rotations
(see Appendix). We simulate the gate fidelity as
a function of the qubit frequency detuning for
variants of the
derivative method, sample method, and
CORPSE pulse (see Figure 2a). We also find that
increasing the gate duration allows the optimizer
to find a more robust pulse
(see Figure 2b). Increasing the order of the derivative
and sample methods requires propagating an additional quantum state
in the augmented state vector, a linear increase
in the computational complexity, but finding higher order
variants of the CORPSE family is prohibitively difficult.

Additionally, we compare the robust control
schemes to a dynamic decoupling pulse
for sequential $R_{x}(\pi/2)$ gates
subject to $T_{\phi}$ noise to demonstrate the applicability
of our techniques to mitigate noise of this type.
We simulate the gate fidelity as a function of time
for variants of the derivative method,
sample method, and B2CORPSE pulse (see Figure 3).
The B2CORPSE pulse is constructed to be robust
to qubit frequency detuning errors and
amplitude scaling errors (see Appendix).
The derivative and sampling methods are employed to
mitigate error due to qubit frequency detuning
($\omega_{q} \gets \omega_{q} + \delta \omega_{q}$)
as well as flux amplitude scaling error
($a \gets a (1 + \delta a)$).

@@

%% AB
\section{Experiment}
We measure $T_{1}$ using the standard experiment
and $T_{2}$ using te Ramsey experiment. We fit with splines
and the data looks like fig. 3 in Helin's paper \cite{zhang2020universal}.
We measure $\omega_{q}$ and $\sigma_{\omega_{q}}$ using X method.

@@

%% AD
\section{Derivative Method Dynamics}
To obtain the dynamics for the derivative of the state $\partial_{x}^{l} \ket{\psi(t)}$
we differentiate the TDSE dynamics \ref{eq:tdse} with respect to the parameter of interest
($x$). Using the fluxonium hamiltonian \ref{eq:hamiltonian} we obtain the derivative of the
state with respect to the qubit frequency ($\omega_{q}$) and the flux amplitude ($a$).
Here we present the dynamics for the second derivative of the state with respect to the
qubit frequency. The case is analogous for the flux amplitude. Both $H$ and $\ket{\psi}$ are functions
of $\omega_{q}$, $a$, and $t$, but we omit the explicit dependence in notation for
brevity.
\begin{equation}
  \begin{aligned}
      \partial_{\omega_{q}} \partial_{t} \ket{\psi} &= \partial_{\omega_{q}} H \ket{\psi}\\
      &= (\partial_{\omega_{q}} H) \ket{\psi} + H (\partial_{\omega_{q}} \ket{\psi})\\
      &= \frac{\sigma_{z}}{2} \ket{\psi} + H (\partial_{\omega_{q}} \ket{\psi})
  \end{aligned}
\end{equation}
\begin{equation}
  \begin{aligned}
    \partial_{\omega_{q}}^{2} \partial_{t} \ket{\psi} &= \partial_{\omega_{q}} (\partial_{\omega_{q}} H \ket{\psi})\\
    &= (\partial_{\omega_{q}}^{2} H) \ket{\psi} + 2 (\partial_{\omega_{q}} H)(\partial_{\omega_{q}} \ket{\psi})\\
    &\quad + H (\partial_{\omega_{q}}^{2} \ket{\psi})\\
    &= \sigma_{z} (\partial_{\omega_{q}} \ket{\psi}) + H (\partial_{\omega_{q}}^{2} \ket{\psi})
  \end{aligned}
\end{equation}
The augmented state vector carries $\partial_{\omega_{q}}^{l} \ket{\psi}$
which appears explicitly in its own dynamics. Due to the dependence of $H$ on $\omega_{q}$ and $a$,
the $l$\textsuperscript{th} state derivative is coupled to the
$l - 1$\textsuperscript{th} state derivative.

>>>>

%% ABSTRACT
\begin{abstract}
  %% - 1 + 2 topic importance
  %% - 1 what I have done
  %% - few sentences about primary results
  The ability to engineer high fidelity operations on quantum processors in the presence of
  calibration errors and decoherence remains the primary challenge requisite to achieving quantum advantage.
  Quantum optimal control (QOC) techniques have proven effective in realizing high fidelity operations,
  but they require exquisite calibration to be performant. In this work we employ robust control techniques
  to achieve gate errors two orders of magnitude lower than dynamic decoupling for parameter
  spreads on the order of $1\%$. 
  Additionally, current QOC techniques mitigate decoherence effects due to longitudinal relaxation by integrating
  Lindblad master equation dynamics, which exponentially increase the computational complexity
  of the optimization algorithm.
  In this work we achieve a factor of $5$ increase in longitudinal relaxation times
  for the gate set of a fluxonium qubit
  using a computationally efficient metric in optimizaiton.
\end{abstract}

@@

%% F1
\begin{figure}[ht]
  \begin{subfigure}{.55\linewidth}
    \includegraphics[width=\linewidth]{assets/f1a.png}
  \end{subfigure}\hfill
  \begin{subfigure}{.4\linewidth}
    \includegraphics[width=\linewidth]{assets/f1b.png}
  \end{subfigure}

  \begin{subfigure}{\linewidth}
    \includegraphics[width=\linewidth]{assets/f1c.png}
  \end{subfigure}
  \caption{
    (a) $T_{1}$ optimized gates (red) and analytic gates (blue).
    (b) $T_{1}$ interpolation function used in optimization. Markers
    denote the time-averaged, absolute amplitude of each gate.
    (c) Master equation simulation with $T_{1}$ dissipation
    over the course of concatenated gate applications.
  }
\end{figure}

%% S3
\section{QOC on the Fluxonium}
(Fluxonium Device) In the following we study
the quantum optimal control problem on the fluxonium qubit.
In the two-level
approximation the system Hamiltonian takes the form
\label{eq:hamiltonian}
\begin{align}
  H/h &= f_{q} \frac{\sigma_{z}}{2} + a(t) \frac{\sigma_{x}}{2}
\end{align}
where $f_{q}$ is the qubit frequency at the flux frustration point,
$a$ is the flux drive amplitude, $h$ is Planck's constant, and $\sigma_{x}, \sigma_{y}$
are Pauli matrices. The flux amplitude $a$ is experimentally
realized by modulating the flux 
threading the device ($\Phi_{\textrm{ext}}$).
The flux amplitude is obtained from the external flux by
$a = 4 \pi \braket{g \lvert \hat{\phi} \rvert e} \rvert_{0.5 \Phi_{0}} E_{L}
\delta \Phi_{\textrm{ext}} / (h \Phi_{0})$
where $\hat{\phi}$ is the phase operator, $E_{L}$ is the characteristic inductance energy, $\Phi_{0}$
is a flux quantum, and
$\delta \Phi_{\textrm{ext}} = \Phi_{\textrm{ext}} - 0.5 \Phi_{0}$ is the flux
offset from the flux frustration point.

@@

%% T1
%% TODO: maybe for appendix.
\begin{table}[ht]
  \begin{tabular}{c | c | c | c}
    & Analytic  & QOC &\\
    Gate & $P_{1}\ (10^{-5})$ & $P_{1}\ (10^{-5})$ & $P_{1\textrm{A}} / P_{1\textrm{Q}}$\\
    \hline
    Z/2 & 5.745 & 1.149 & 5.000\\
    Y/2 & 5.253 & 1.157 & 4.540\\
    X/2 & 16.251 & 6.946 & 2.340\\
  \end{tabular}
  \caption{Probability of longitudinal relaxation for each gate
    evaluated at the gate's duration.}
\end{table}

%% F2
\begin{figure}[ht]
  \begin{subfigure}{.55\linewidth}
    \includegraphics[width=\linewidth]{assets/f2a.png}
  \end{subfigure}\hfill
  \begin{subfigure}{.4\linewidth}
    \includegraphics[width=\linewidth]{assets/f2b.png}
  \end{subfigure}

@@

  \caption{
    (a) X/2 gates robust to qubit frequency detunings constructed using the
    analytic, (2, 4)-point sampling, and (2\textsuperscript{nd},
    3\textsuperscript{rd})-order derivative methods.
    (b) Gate error as a function of the detuning from the nominal
    qubit frequency at a fixed gate duration, see (a), and at the best
    gate duration for each method chosen from a sweep, see (c).
    (c) Gate error at a one-percent detuning from the nominal qubit frequency as
    a function of the gate duration set in optimization.
  }
\end{figure}

@@

  \caption{
    (a) $X/2$ gates robust to flux offsets constructed using the analytic,
    (2, 4)-point sampling, and (2, 3)-derivative methods at a gate duration
    of $56.8 (ns)$. (b) Mean gate error for 10 samples of 1/$f$ flux noise over 5000
    gate applications.
    %% (c) Stochastic 1/$f$ flux noise.
  }
\end{figure}

%% S5
\section{Robustness to Parameter Deviations}
(Strategy) For the fluxonium, pure dephasing acts
along the flux axis. Low frequency 1/$f$ noise,
Johnson-Nyquist current noise present in the flux bias
lines, and temperature dependent gain fluctuations
may cause the flux amplitude to deviate from its
desired value by an amount $\delta a$. Additionally,
residual calibration errors and fluctuations may cause
the qubit frequency to deviate from its measured value
by an amount $\delta f_{q}$.

@@

The second scheme we analyze is the sampling method,
which is well studied in the robust control literature
\cite{manchester2016derivative} \cite{tronarp2016sigma}.
We add additional
states that follow deviant dynamics to the augmented
state vector. The parameters of interest
are altered by a small deviation, typically chosen
to represent the distribution of the parameter.
For example, in the 2-point sampling method we
propagate $\ket{\psi^{\pm, I}(t)}$ where
$f_{q}^{\pm} = f_{q} \pm \sigma_{f_{q}}$,
$f_{q}^{I} = f_{q}$.
The difference between the final states and the target
state is penalized, resulting in quadratic costs at
each knot point
${\lvert \ket{\psi^{\pm, I}_{k}} - \ket{\psi_{f}} \rvert}^{2}$

(Results) We compare the robust control
schemes to a dynamic decoupling pulse
for a single $R_{x}(\pi/2)$ gate
subject to qubit frequency detuning
($f_{q} \gets f_{q} + \delta f_{q}$)
to demonstrate the applicability of our techniques to mitigate
system parameter deviations.
We use a compensating for off-resonance error
(CORPSE) dynamic decoupling pulse that is optimized
to mitigate first-order error arising due to the detuning.
The pulse is composed of
three rotations about the $\hat{x}$, $-\hat{x}$, and $\hat{x}$ axes
of the Bloch sphere
designed to compensate for erroneous rotations
(see Appendix). We simulate the gate fidelity as
a function of the qubit frequency detuning for
variants of the
derivative method, sample method, and
CORPSE pulse (see Figure 2a). We also find that
increasing the gate duration allows the optimizer
to find a more robust pulse
(see Figure 2b). Increasing the order of the derivative
and sample methods requires propagating an additional quantum state
in the augmented state vector, a linear increase
in the computational complexity, but finding higher order
variants of the CORPSE family is prohibitively difficult.

Additionally, we compare the robust control
schemes to a dynamic decoupling pulse
for sequential $R_{x}(\pi/2)$ gates
subject to $T_{\phi}$ noise to demonstrate the applicability
of our techniques to mitigate noise of this type.
We simulate the gate fidelity as a function of time
for variants of the derivative method,
sample method, and B2CORPSE pulse (see Figure 3).
The B2CORPSE pulse is constructed to be robust
to qubit frequency detuning errors and
amplitude scaling errors (see Appendix).
The derivative and sampling methods are employed to
mitigate error due to qubit frequency detuning
($f_{q} \gets f_{q} + \delta f_{q}$)
as well as flux amplitude scaling error
($a \gets a (1 + \delta a)$).

@@

%% AB
\section{Experiment}
We measure $T_{1}$ using the standard experiment
and $T_{2}$ using te Ramsey experiment. We fit with splines
and the data looks like fig. 3 in Helin's paper \cite{zhang2020universal}.
We measure $f_{q}$ and $\sigma_{f_{q}}$ using X method.

@@

%% AD
\section{Derivative Method Dynamics}
To obtain the dynamics for the derivative of the state $\partial_{x}^{l} \ket{\psi(t)}$
we differentiate the TDSE dynamics \ref{eq:tdse} with respect to the parameter of interest
($x$). Using the fluxonium hamiltonian \ref{eq:hamiltonian} we obtain the derivative of the
state with respect to the qubit frequency ($f_{q}$) and the flux amplitude ($a$).
Here we present the dynamics for the second derivative of the state with respect to the
qubit frequency. The case is analogous for the flux amplitude. Both $H$ and $\ket{\psi}$ are functions
of $f_{q}$, $a$, and $t$, but we omit the explicit dependence in notation for
brevity.
\begin{equation}
  \begin{aligned}
      \partial_{f_{q}} \partial_{t} \ket{\psi} &= \partial_{f_{q}} H \ket{\psi}\\
      &= (\partial_{f_{q}} H) \ket{\psi} + H (\partial_{f_{q}} \ket{\psi})\\
      &= \frac{\sigma_{z}}{2} \ket{\psi} + H (\partial_{f_{q}} \ket{\psi})
  \end{aligned}
\end{equation}
\begin{equation}
  \begin{aligned}
    \partial_{f_{q}}^{2} \partial_{t} \ket{\psi} &= \partial_{f_{q}} (\partial_{f_{q}} H \ket{\psi})\\
    &= (\partial_{f_{q}}^{2} H) \ket{\psi} + 2 (\partial_{f_{q}} H)(\partial_{f_{q}} \ket{\psi})\\
    &\quad + H (\partial_{f_{q}}^{2} \ket{\psi})\\
    &= \sigma_{z} (\partial_{f_{q}} \ket{\psi}) + H (\partial_{f_{q}}^{2} \ket{\psi})
  \end{aligned}
\end{equation}
The augmented state vector carries $\partial_{f_{q}}^{l} \ket{\psi}$
which appears explicitly in its own dynamics. Due to the dependence of $H$ on $f_{q}$ and $a$,
the $l$\textsuperscript{th} state derivative is coupled to the
$l - 1$\textsuperscript{th} state derivative.
